---
title: "Pseudotime project clean"
output: html_document
date: "2026-04-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Run this to install packages:
{
if (!require("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

## Install necessary packages
BiocManager::install(c('BiocGenerics', 'DelayedArray', 'DelayedMatrixStats',
                       'limma', 'lme4', 'S4Vectors', 'SingleCellExperiment',
                       'SummarizedExperiment', 'batchelor', 'HDF5Array',
                       'ggrastr', 'Seurat', 'dplyr', 'ggplot2','scRNAseq',
                       'satijalab/seurat-wrappers'), dependencies = TRUE)

install.packages("devtools")
remotes::install_github("bnprks/BPCells/r")
devtools::install_github('cole-trapnell-lab/monocle3')


remotes::install_github("LTLA/scRNAseq")

remove.packages("AnnotationHub")
BiocManager::install("AnnotationHub", force = TRUE)
## Delete the local AnnotationHub cache entirely
ah_cache <- tools::R_user_dir("AnnotationHub", which = "cache")
unlink(ah_cache, recursive = TRUE)

## Same for ExperimentHub
eh_cache <- tools::R_user_dir("ExperimentHub", which = "cache")
unlink(eh_cache, recursive = TRUE)

# Project

```{r message = FALSE}
#This step can take a bit longer
rm(list = ls())

# Restart R, then retry
library(scRNAseq)
sce <- NestorowaHSCData()

# Load required libraries
library(monocle3)
library(Seurat)
library(SeuratWrappers)
library(ggplot2)
library(SingleCellExperiment)
library(dplyr) 

#Create seurat object
seurat <- CreateSeuratObject(counts = counts(sce), meta.data = as.data.frame(colData(sce)))

#Remove no longer necessary object
rm(list = c("ah_cache", "eh_cache", "sce"))
```

# Seurat Object Processing
```{r}
seurat <- NormalizeData(seurat)
seurat <- FindVariableFeatures(seurat, nfeatures = 2000)
all_genes <- rownames(seurat)
seurat <- ScaleData(seurat, features = all_genes)
seurat <- RunPCA(seurat, verbose = TRUE)
seurat <- FindNeighbors(seurat, dims = 1:20, verbose = T)
seurat <- FindClusters(seurat)
seurat <- RunUMAP(seurat, dims = 1:20)
```

# Conversion to monocle3 cds
```{r}
# Convert Seurat object to Monocle3 cell_data_set
cds <- as.cell_data_set(seurat)

#Add gene names to row data
rowData(cds)$gene_short_name <- rownames(seurat)
```

**Topic 3 (Groups 5 & 6): Gene expression dynamics along pseudotime**
Identify genes whose expression changes along differentiation trajectories.
Suggested steps:
- Select a subset of cells (e.g. one lineage or branch)
- Use graph_test() to identify genes associated with pseudotime
- Select the top ~5 genes based on statistical significance
- Visualize their expression along pseudotime (e.g. plot_genes_in_pseudotime())
Questions to consider:
- Do genes show clear trends (e.g. increasing, decreasing, transient expression)?
- Are there differences between branches or cell types?
- Are any of these genes known to play a role in hematopoiesis?

```{r}
#preprocessing
cds <- preprocess_cds(cds, num_dim = 100)
#reduce dimensions
cds <- reduce_dimension(
  cds,
  reduction_method = "UMAP",
  umap.min_dist = 0.1,      
  umap.n_neighbors = 20         
)
#cluster cells
cds <- cluster_cells(cds, resolution = 1e-1, k = 10)
cds <- learn_graph(cds)

plot_cells(cds, color_cells_by = "gate",
           label_cell_groups= FALSE,
           label_groups_by_cluster = FALSE, 
           label_leaves = FALSE,
           cell_size = 0.5) +

plot_cells(cds, color_cells_by = "partition",
           label_cell_groups= FALSE,
           label_groups_by_cluster = FALSE, 
           label_leaves = FALSE,
           cell_size = 0.5) +

plot_cells(cds, color_cells_by = "broad",
           label_cell_groups= FALSE,
           label_groups_by_cluster = FALSE, 
           label_leaves = FALSE,
           cell_size = 0.5)
```

# Subsetting to celltypes
```{r}
sub_cds <- cds[,colData(cds)$broad %in% c("LTHSC","MPP","CMP","MEP")]

sub_cds <- preprocess_cds(sub_cds, num_dim = 100)
sub_cds <- reduce_dimension(sub_cds, reduction_method = "UMAP")
sub_cds <- cluster_cells(sub_cds, reduction_method = "UMAP")
sub_cds <- learn_graph(sub_cds)

plot_cells(sub_cds, color_cells_by = "broad",
           label_cell_groups= FALSE,
           label_groups_by_cluster = FALSE, 
           label_leaves = FALSE,
           cell_size = 0.5,
           reduction_method = "UMAP")
```

# Ordering Pseudotime
```{r}
sub_cds <- order_cells(sub_cds)
```

# Identifying marker genes
```{r}
#genes_out_new <- graph_test(sub_cds, neighbor_graph="principal_graph", cores=16)
genes_out_new <- readRDS('/Users/Shared/GitHub/Focus I - RNA Seq Practical/Results/genes_out_new.RDS')
saveRDS(genes_out_new, file = '/Users/Shared/GitHub/Focus I - RNA Seq Practical/Results/genes_out_new.RDS')

hvg <- VariableFeatures(seurat) 


TopGenes <- genes_out_new %>%
  filter(status == "OK") %>%
  filter(q_value > 0) %>%
  arrange(q_value) %>%
  slice_head(n=5) %>%
  pull(gene_short_name)

TopGenes
```

# Plotting marker genes
```{r}
gene_sub <- sub_cds[rowData(sub_cds)$gene_short_name %in% TopGenes,]

plot_cells(sub_cds, genes = TopGenes,
           label_cell_groups= FALSE,
           label_groups_by_cluster = FALSE, 
           label_leaves = FALSE,
           cell_size = 0.5,
           reduction_method = "UMAP")
  
plot_genes_in_pseudotime(gene_sub)

```

