---
title: "Pseudotime project clean"
output: html_document
date: "2026-04-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Run this to install packages:
{
if (!require("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

## Install necessary packages
BiocManager::install(c('BiocGenerics', 'DelayedArray', 'DelayedMatrixStats',
                       'limma', 'lme4', 'S4Vectors', 'SingleCellExperiment',
                       'SummarizedExperiment', 'batchelor', 'HDF5Array',
                       'ggrastr', 'Seurat', 'dplyr', 'ggplot2','scRNAseq',
                       'satijalab/seurat-wrappers'), dependencies = TRUE)

install.packages("devtools")
remotes::install_github("bnprks/BPCells/r")
devtools::install_github('cole-trapnell-lab/monocle3')


remotes::install_github("LTLA/scRNAseq")

remove.packages("AnnotationHub")
BiocManager::install("AnnotationHub", force = TRUE)
## Delete the local AnnotationHub cache entirely
ah_cache <- tools::R_user_dir("AnnotationHub", which = "cache")
unlink(ah_cache, recursive = TRUE)

## Same for ExperimentHub
eh_cache <- tools::R_user_dir("ExperimentHub", which = "cache")
unlink(eh_cache, recursive = TRUE)

# Project

```{r message = FALSE}
#This step can take a bit longer
rm(list = ls())

# Restart R, then retry
library(scRNAseq)
sce <- NestorowaHSCData()

# Load required libraries
library(monocle3)
library(Seurat)
library(SeuratWrappers)
library(ggplot2)
library(SingleCellExperiment)
library(dplyr) 

#Create seurat object
seurat <- CreateSeuratObject(counts = counts(sce), meta.data = as.data.frame(colData(sce)))

#Remove no longer necessary object
rm(list = c("ah_cache", "eh_cache", "sce"))
```

# Seurat Object Processing
```{r}
seurat <- NormalizeData(seurat)
seurat <- FindVariableFeatures(seurat, nfeatures = 2000)
seurat <- ScaleData(seurat, features = all_genes)
seurat <- RunPCA(seurat, verbose = TRUE)
seurat <- FindNeighbors(seurat, dims = 1:20, verbose = T)
seurat <- FindClusters(seurat)
seurat <- RunUMAP(seurat, dims = 1:20)
```

# Conversion to 
```{r}

```


